---
title: "penguin_analysis"
output:
  pdf_document: default
  html_document: default
  word_document: default
date: "2023-12-04"
---

### Introduction

In this rmarkdown project, I plan to explore and statistically analyse the PalmerPenguins dataset. Specifically, I will be exploring and analysing the relationship between body mass and flipper length for each penguin species.Studying flipper length and its relationship to body mass could have crucial conservation implications. As climate change increases, this may impact the fitness of penguin populations, and monitoring the relationship between different physical variables will benefit assessing the health and adaptability of penguin populations.

In order to study the relationship between body mass and flipper length, I will load the raw data from the PalmerPenguin dataset, and evaluate if it needs to be cleaned. Then I will provide an exploratory figure to visualise the relationship between body mass and flipper length for each species, providing hypotheses on the relationship. Following this I will statistically test the data and provide a results figure from my statistical analysis. Finally, I discuss these results, and then provide my concluding remarks. 


```{r data exploration, echo=TRUE, warning=FALSE, message=FALSE}
#if project out of synch with renv, run renv::restore() in console
#unhash below to install packages if not already installed : 
#install.packages(c("ggplot2", "palmerpenguins", "janitor", "dplyr", "ragg", "svglite"))

# Load the packages:
library(ggplot2)
library(palmerpenguins)
library(janitor)
library(dplyr)
library(ragg)
library(svglite)

# Load the function definitions
source("functions/cleaning.r")
source("functions/plotting.r")

#Save the raw data as a csv into the "data_raw" folder : 
write.csv(penguins_raw, "data_raw/penguins_raw.csv")

#Load the raw penguin data from the previously saved csv : 
penguins_raw <- read.csv("data_raw/penguins_raw.csv")

#Check the raw data :  

head(penguins_raw)

#species names quite long - need to clean them to shorten for easier analysis

#checking column names in raw data : 
names(penguins_raw)

#names are not human or computer readable - need to clean them


#Clean the raw data calling functions from 'source('functions/cleaning.r') : 

penguins_clean <- penguins_raw %>% 
  clean_column_names %>% 
  shorten_species %>% 
  remove_empty_columns_rows

#Save the clean data : 

write.csv(penguins_clean, "data_clean/penguins_clean.csv")

#check the clean data : 

names(penguins_clean)
head(penguins_clean)

##I am exploring the relationship between body mass and flipper length 
#Subset clean data to only include species, body mass and flipper length : 
#Remove NA's :
#Call the cleaning function "remove_empty_mass_flippers" to do so :

penguins_mass_flippers <- remove_empty_mass_flippers(penguins_clean)

#check the new dataset including only species, body mass, and flipper length

head(penguins_mass_flippers)

#Call the "plot_mass_flipper_figure" function 
#from 'source("functions/plotting.r")' to plot scatterplot : 

mass_flipper_scatterplot <- plot_mass_flipper_figure(penguins_mass_flippers)

mass_flipper_scatterplot

##Saving the scatterplot as an image in figures folder steps : 

#calling 'save_mass_flipper_plot_png' ('source(functions/plotting.r)')
#saving scatterplot as a png in figures folder
save_mass_flipper_plot_png(penguins_mass_flippers, 
                      "figures/mass_flipper_scatterplot.png", 
                      size = 10, res = 600, scaling=0.7)

#calling 'save_mass_flipper_plot_svg' ('source(functions/plotting.r'))
#saving scatterplot as an svg in figures folder 
save_mass_flipper_plot_svg(penguins_mass_flippers, 
                      "figures/mass_flipper_scatterplot.svg", 
                      size = 10, scaling = 1)

```

### Hypothesis

From the output of this plot, I hypothesise that there is a positive correlation between body mass and flipper length for all three penguin species as it appears from individual data points on the scatter plot that as body mass increases, so does flipper length. I also hypothesise that this positive correlation is statistically significant. There appears to be a lot of variability in the data points for each species, but less so for Gentoo and Chinstrap species. Therefore, I also hypothesise that body mass will account for more variation in the data for these two species, but will not account for all data variation for any species. 

For the rest of this rmarkdown project, I will be statistically testing the null hypothesis that there is no significantly positive correlation between flipper length and body mass for each of the three penguin species. The alternative hypothesis I will be testing is that there is indeed a statistically significant positive correlation between these two variables for each of the three penguin species.

### Statistical Methods

In order to test my hypotheses, I will use a linear regression analyss to determine if there is a positive correlative relationship between body mass and flipper length for each of the three species. I will also use a linear regression analysis to obtain a correlation coefficient and determine if there is a positive correlation between the two variables. I will also obtain p-values to determine if the results are significant. If the p value is below 0.5, then this indicates that there is at least one coefficient in my model that is not zero, providing support to reject the null hypothesis, and accept the alternative hypothesis. I can also obtain an R-squared value which will provide information as to how much variability in the dependent variable (flipper length) can be explained by the independent variable (body mass) in my model. Additionally, by using a linear regression analysis, I can fit a regression line onto the above scatter plot to visualise if the relationship between body mass and flipper length is positive or negative. 

```{r statistics, echo=TRUE,eval=TRUE, warning=FALSE, message=FALSE} 

#Subset the penguins_mass_flippers data for each species : 
adelie_mass_flippers_data <- subset(penguins_mass_flippers, species == "Adelie")
chinstrap_mass_flippers_data <- subset(penguins_mass_flippers, species == "Chinstrap")
gentoo_mass_flippers_data <- subset(penguins_mass_flippers, species == "Gentoo")

##do a linear regression analysis on the subset species data : 

#Adelie : 

lm_model_adelie <- lm(flipper_length_mm ~ body_mass_g, data = adelie_mass_flippers_data)

#Chinstrap : 

lm_model_chinstrap <- lm(flipper_length_mm ~ body_mass_g, data = chinstrap_mass_flippers_data)

#Gentoo : 

lm_model_gentoo <- lm(flipper_length_mm ~ body_mass_g, data = gentoo_mass_flippers_data)

##Get a summary of the linear regression model for each species : 

#Adelie : 
summary_adelie_lm_model <- summary(lm_model_adelie)
summary_adelie_lm_model

#Chinstrap : 

summary_chinstrap_lm_model <- summary(lm_model_chinstrap)
summary_chinstrap_lm_model

#Gentoo

summary_gentoo_lm_model <- summary(lm_model_gentoo)
summary_gentoo_lm_model

```

### Results & Discussion

From the summary of the linear regression models for each species, I will add a regression line and confidence intervals onto my previous scatter plot looking at the association between body mass and flipper length. 

```{r Plotting Results, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}

#Call 'results_mass_flipper_figure' function 
#from 'source("functions/plotting.r")' :

mass_flipper_results_plot <- results_mass_flipper_figure(penguins_mass_flippers)
mass_flipper_results_plot

##Saving the results scatterplot as an image in figures folder steps : 

#calling 'save_mass_flipper_plot_png' ('source(functions/plotting.r)')
#saving scatterplot as a png in figures folder
save_mass_flipper_results_png(penguins_mass_flippers, 
                      "figures/mass_flipper_resultsplot.png", 
                      size = 10, res = 600, scaling=0.7)


#calling 'save_mass_flipper_plot_svg' ('source(functions/plotting.r'))
#saving scatterplot as an svg in figures folder 
save_mass_flipper_results_svg(penguins_mass_flippers, 
                      "figures/mass_flipper_resultsplot.svg", 
                      size = 10, scaling = 1)


```

Results from the linear regression model support my hypothesis that there is an association between body mass and flipper length for all species, and that it is a positive correlation. 

Focussing on Adelie penguin species, the linear regression model provides a p-value < 0.05, of 1.343e-09. This is highly significant, leading me to reject the null hypothesis, providing strong evidence a relationship exists between body mass and flipper length within this species. A highly significant correlation coefficient was found with a value of 6.667e-03, with a p-value of 1.34e-09 (significance level <0.0001). As the correlation coeffcient is greater than 0, this demonstrates that there is a positive correlation between body mass and flipper length in the Adelie species. However, the correlation coefficient is very close to 0, suggesting that this is not a strong positive relationship. An R-squared value of 0.2192 was found, suggesting that body mass only explains 21.92% of variation in the flipper length data. Therefore, I suggest there are other factors other than body mass which better explain the variation of flipper length data, and the linear model does not appear to be the most appropriate model to fit this data. 

The linear regression model for Chinstrap species also appears to be highly significant, with a p value < 0.05 of 3.748e-09. Again, I can reject the null hypothesis and this provides strong evidence of a relationship between body mass and flipper length within this species. A highly significant correlation coefficient of 1.191e-02 was also found, with a p-value of 3.75e-09 (significance level <0.0001). This suggests a positive correlation between body mass and flipper length as it is greater than 0. This correlation coefficient is also higher than that found for Adelie penguin species, suggesting that body mass influences flipper length more so in this species. This is supportive of my earlier hypothesis that body mass would explain more variation in Chinstrap species. The correlation coefficient, however, is still close to 0, so the positive relationship is a weak one. An R-squared value of 0.4116 was found, suggesting that body mass explains 41.26% of the variation in flipper length data, which is more so than what was explained for in Adelie species. This further supports that body mass influences flipper length more so in chinstrap species than in Adelie species, but the majority of the variation in flipper length data appears to be still better explained by alternative variables. 

Finally looking at Gentoo penguins, a p-value for the linear regression model of < 2.2e-16 was found. As p<0.05, I can reject the null hypothesis. The p-value also provides support that a relationship exists between body mass and flipper length within this species. A correlation coefficient of 9.039e-03 was found, and was highly significant, with a p value of <2e-16 (significance level <0.0001). As the correlation coefficient is greater than 0, this provides evidence that there is a positive correlation between body mass and flipper length for this species. The correlation coefficient is also greater than that for Adelie penguins, but smaller than that for Chinstrap penguins. This suggests that overall, body mass influences flipper length more so for Gentoo and Chinstrap penguins than it does for Adelie penguins. This supports my hypothesis that body mass will account for more variation in flipper length in Gentoo ad Chinstrap penguins. It should be noted that the correlation coefficient for Gentoo penguin data is still close to 0, also suggesting a weak positive relationship. An R-squared of 0.4937 was found so body mass explains 49.37% of variation in flipper length data, suggesting body mass influences flipper length a lot more compared to the previous two species than other variables may have. Despite this, ~50% of variation is still unexplained, and similar to the other two species, variation in flipper length in Gentoo penguins could be explained by other variables. This supports my hypothesis that body mass would not be able to explain all the variation in the flipper length data, suggesting there are other contributing variables. 

Overall, the linear regression model provides strong evidence that there is a positive, correlative association between body mass and flipper length in all three species, but to varying degrees. One reason as to why flipper length may increase with body mass in all three species is due to biological scaling. In many biological organisms, there are scaling relationships between different body parts, so as an organism grows, different body parts may exhibit proportional changes. Consequently this could result in a positive correlation between body mass and flipper length in some species due to scaling relationships. A second reason as to why flipper length increases with body mass could be due to functional adaptations. In penguin species, flipper length is often related to swimming ability in their aquatic environments, so a larger body mass may require longer flippers for more efficient swimming. 

It is clear however that there are other factors influencing flipper length other than body mass, as body mass alone does not explain most of the variation in the data.This could be due to ecological niche and the species may be under different selective pressures favouring certain body proportions. It may be that Adelie penguins do not have to rely so heavily on hunting or escaping predators, so there may be no reason to have longer flippers with increased body mass, hence why body mass explains the lowest amount of data variation. I suggest in future an appropriate analysis to test this would be an ancova on the data of flipper length of the different species, and the islands they come from. 

These different islands may also have different environments, which could also affect penguin species. Penguin species inhabit diverse climates and geographic locations, ranging from cold climates in Antarctica, to more temperate and tropical climates in Australia and the Galapagos (respectfully). Understanding the environmental factors that determine flipper length reveals how different penguin species evolved to their aquatic and terrestrial environments, which provides important biological insight into evolutionary adaptations. This could also impact conservation. Dramatic changes in environmental conditions, such as shifts in prey availability and climate change, may impact the selective pressures on penguin populations, thus their flipper length. Monitoring flipper length of different penguin species, and variation in flipper length, can provide early indicators of how penguins are responding to environmental changes. 

In some penguin species sexual dimorphism is present, so males and females present differences in size and other physical characteristics. There may be sexual dimorphism in body size or flipper length which could influence the relationship between these traits, and could be why different penguin species show different relationships between the two variables. Males and females play different reproductive roles, so there may be different evolutionary pressures leading differences in male and female body size and flipper length for different species. Females may be primarily responsible for incubating eggs, such as in Adelie penguins, while males forage, so their physical characteristics may be adapted to these roles. Therefore, males and females may not present the same relationship between body mass and flipper length. An appropriate analysis to test this in the future would be a linear regression or an ancova to assess whether there are significant differences in the relationship between body size and flipper length based on sex for different species. Understanding flipper length, and how it relates to sexual dimorphism provides important biological insight into how reproductive strategies can shape a penguin species' morphology and diversity within it. This also is important to enhancing biological understanding of how different penguin species adapted to their environments as a result of this. 

### Conclusion

In conclusion, body mass is associated with penguin flipper length across all three species in the PenguinPalmer dataset. There is a positive relationship between these two variables for all three species, so as body mass increases, flipper length also tends to increase. This may be due to biological allometric scaling of body parts, and functional adaptations in penguin species, perhaps associated with swimming. This information is supportive of my hypothesis that there is a significant positive correlation between body mass and flipper length. 

Although there is a positive correlation, there is a weak association between body mass and flipper length. Additionally, body mass fails to explain most of the variation in flipper length data across all species, though body mass explained more variation in Chinstrap and Gentoo penguin species. This is supportive of my hypothesis that body mass will account for more variation in the data for these two species, but will not account for all data variation for any species. This suggests there are other contributing variables that affect flipper length, possible variables may be ecological niche and sexual selection, but these should be tested with other data from the PalmerPenguins dataset such as the island species come from, and sex within species.



